---
title: "sciModelFlowR 0.3.0: Probabilistic Machine Learning"
vignette: >
  %\VignetteIndexEntry{sciModelFlowR 0.3.0: Probabilistic Machine Learning}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
execute:
  echo: true
  warning: true
  message: false
---

# Probabilistic prediction as a scientific contract

Version 0.3.0 introduces the first stable probabilistic layer in `sciModelFlowR`. The objective is not to label every uncertain output a “distribution.” The package distinguishes point predictions, class probabilities, empirical samples, quantiles, parametric distributions, and interval-only objects. Operations that are not mathematically available for a representation must fail rather than fabricate a density or variance.

The central rule is:

**Name the predictive object first, then choose scores and uncertainty statements that are valid for that object.**

## 1. Learning objectives

The reader should be able to distinguish a class-probability vector from a continuous predictive distribution, explain why CRPS and log loss answer different questions, calculate calibration error without using final test labels to recalibrate the model, understand sharpness versus calibration, request only distribution operations supported by the fitted representation, and report the source and target of uncertainty explicitly.

## 2. The stable `PredictionDistribution`

A prediction distribution is created with:

```{r}
#| eval: false
smf_prediction_distribution()
```

The supported representations in 0.3.0 are `point`, `class_probabilities`, `samples`, `quantiles`, `parametric`, and `interval`. The payload differs by representation. The stable object records what the representation is and which operations it can support.

Representative operations are:

```{r}
#| eval: false
smf_dist_mean()
smf_dist_median()
smf_dist_variance()
smf_dist_sd()
smf_dist_quantile()
smf_dist_interval()
smf_dist_sample()
smf_dist_log_prob()
smf_dist_cdf()
```

A class-probability matrix supports log probability for observed class labels and calibration diagnostics. It does not expose a meaningful numeric predictive mean merely because the class labels can be coerced to integers.

## 3. A parametric Gaussian example

```{r}
#| eval: false
pd <- smf_prediction_distribution(
  "parametric",
  payload = list(
    distribution = "normal",
    mean = c(50, 55, 60),
    sd = c(3, 4, 5)
  )
)

smf_dist_mean(pd)
smf_dist_quantile(pd, c(0.025, 0.5, 0.975))
smf_dist_interval(pd, level = 0.95)
```

For a Normal distribution, the mean, variance, quantiles, density, CDF, and random sampling are well defined. If a later model only returns two interval bounds, however, those bounds do not identify a unique density. The package therefore does not infer an arbitrary Normal distribution from an interval.

## 4. Class probabilities

A binary or multiclass classifier in 0.3.0 produces named probability columns. The package validates three basic properties: probabilities are finite and remain in the unit interval, each row sums to one within numerical tolerance, and the observed class labels can be mapped to columns.

```{r}
#| eval: false
b <- smf_load_dataset("gold_binary_calibration")
b$event <- factor(ifelse(b$event == 1, "yes", "no"), levels = c("no", "yes"))

binary_spec <- smf_experiment_spec(
  task = smf_task_spec("binary", "event", positive_label = "yes"),
  data = smf_data_spec("event", c("x1", "x2"), id_column = "obs_id"),
  design = smf_design_spec(id_column = "obs_id"),
  preprocessing = smf_preprocess_spec(center = TRUE, scale = TRUE),
  resampling = smf_resampling_spec("stratified_holdout", train_prop = .75, strata = "event"),
  model = smf_model_spec("logistic_regression", "stats"),
  metrics = list(smf_metric_spec("log_loss"), smf_metric_spec("brier"))
)

binary_fit <- smf_fit_experiment(binary_spec, b)
pd <- smf_predict_distribution(binary_fit)
```

The observed-class log probability can then be computed through the distribution contract rather than by reaching into a backend object.

## 5. Log loss

For classification, log loss rewards assigning high probability to the event that actually occurs and strongly penalizes confident errors. It therefore detects differences that accuracy cannot see. A classifier assigning 0.51 to the correct class and another assigning 0.99 receive the same hard-label accuracy but different log losses.

The scientific caution is that log loss can be dominated by a small number of extremely confident mistakes. That is not necessarily a defect; it reflects the severe cost of claiming near certainty and being wrong. Inspect such cases rather than hiding them.

## 6. Brier score

The Brier score measures squared error between class indicators and class probabilities. For binary outcomes it is directly analogous to mean squared probability error. For multiclass outcomes, the package computes the row-wise squared discrepancy across all classes and averages it.

Brier score combines calibration and discrimination effects. A good Brier value is useful, but it should not replace a calibration plot or class-specific diagnostic when the intended use depends on trustworthy probabilities.

## 7. CRPS for continuous predictive distributions

CRPS compares an entire predictive CDF with an observed continuous outcome. Version 0.3.0 implements the closed-form Gaussian CRPS for a Normal parametric `PredictionDistribution`. This creates a validated foundation for later distributional models without pretending that every deterministic regression supplies a continuous predictive distribution.

A deterministic point predictor can be scored with RMSE or MAE. It should not receive a “CRPS” produced by assuming an undocumented variance.

## 8. Interval coverage and width

Coverage asks whether the observed value lies within an interval. Width describes how concentrated the interval is. A method can obtain very high coverage by returning excessively broad intervals. Therefore coverage and width must be read together.

`smf_evaluate_probabilistic()` can calculate empirical coverage, mean width, and interval score when the distribution supports quantile extraction. The nominal level should always be reported alongside empirical coverage.

## 9. Interval score

The interval score rewards narrow intervals that contain the observation and penalizes misses in proportion to how far the observation lies outside the bounds. It is a proper scoring rule for central prediction intervals at a specified level.

The interval score is not a replacement for CRPS when a full predictive distribution is available, but it is often useful when a scientific workflow naturally communicates selected uncertainty levels.

## 10. Expected calibration error

For classification, `smf_calibration_report()` bins predicted probabilities and compares the mean probability with observed frequency. The weighted absolute discrepancy is summarized as expected calibration error (ECE).

```{r}
#| eval: false
cal <- smf_calibration_report(
  truth = binary_fit@prediction@truth,
  probabilities = binary_fit@prediction@probabilities,
  bins = 10L
)

cal$ece
cal$table
```

ECE depends on binning. It is therefore a diagnostic summary, not an intrinsic property of a model independent of analyst choices. Report the bin definition or use a calibration curve alongside the number.

## 11. Calibration and discrimination are different

A classifier may rank cases very well but produce probabilities that are systematically too extreme or too conservative. ROC-AUC can remain high while probability calibration is poor. Conversely, a model that predicts the event prevalence for every observation can be well calibrated in a weak aggregate sense but discriminate poorly.

A scientific probability model should be evaluated on both axes. If probabilities drive risk decisions, expected losses, or resource allocation, calibration may be as important as ranking.

## 12. Sharpness

Sharpness describes the concentration of predictive distributions without reference to the observations. A calibrated forecast that is sharper is generally more informative than an equally calibrated but diffuse forecast. Sharpness alone, however, can be maximized by an overconfident model.

The package treats sharpness as a companion to calibration and proper scores, not as an optimization target in isolation.

## 13. PIT diagnostics

Probability integral transform diagnostics are meaningful when a continuous predictive CDF is available. They are not appropriate for a deterministic point estimate or a class-probability vector. Later distributional and Bayesian modules will extend PIT support after the underlying distributions are validated.

This illustrates why `PredictionDistribution` carries representation and capabilities rather than merely a numeric array called “uncertainty.”

## 14. Empirical sample distributions

Bootstrap predictions, ensembles, posterior predictive draws, and MC-dropout samples can all be represented as empirical samples. Their scientific interpretations are not identical. An empirical matrix of samples only says what operations can be performed computationally; the `UncertaintyDescriptor` states where those samples came from and what they target.

Bootstrap sampling variation must not be relabeled posterior uncertainty. Deep-ensemble variation must not automatically be claimed to identify epistemic uncertainty. The package intentionally separates computational representation from inferential interpretation.

## 15. Quantile-only predictions

Some models estimate selected conditional quantiles without providing a likelihood. The quantile representation can support interpolation, selected intervals, coverage, and interval scores. It does not automatically support log density or CDF evaluation outside the fitted quantile grid.

This distinction will be important when quantile boosting and conformalized quantile regression are added in later versions.

## 16. The heteroscedastic Gold dataset

`gold_heteroscedastic_regression` was designed so the conditional mean and conditional variance behave differently across the predictor space. A point model can approximate the mean and still fail to describe changing predictive spread.

Use this fixture to teach why RMSE alone cannot establish quality of uncertainty. A later probabilistic model should be judged by both mean performance and distributional scores, including whether interval width changes where the generator variance changes.

## 17. Scientifically inappropriate workflow

An invalid workflow fits a deterministic regression, calculates its residual standard deviation using the full dataset, attaches `mean ± 1.96 * sd` to every future prediction, and calls the result a calibrated 95% predictive distribution. The procedure leaks test residuals, assumes constant variance, ignores parameter uncertainty, and silently imposes Normality.

The correct response is not to find a more impressive name for the interval. The correct response is to fit or construct an uncertainty method whose assumptions and calibration data are explicit.

## 18. Probability scores and prevalence

Rare-event classification deserves special care. Accuracy can be dominated by the majority class. ROC-AUC can remain visually strong in problems where precision is poor because false positives accumulate against a large negative population. PR-oriented metrics and probability calibration can reveal this behavior.

The `gold_multiclass_imbalanced` fixture provides a controlled setting for these comparisons.

## 19. Probability scale versus decision scale

A probability is not yet a decision. A 0.30 pest-risk estimate can trigger action in one context and no action in another depending on false-negative cost, intervention cost, prevalence, and available capacity. Threshold choice belongs to an explicit decision rule and should be tuned on validation/calibration evidence, not on the final test set.

Keeping probability estimation separate from thresholding allows the same model to be re-evaluated under a different operational objective without retraining.

## 20. Reporting probabilistic predictions

Report the predictive representation, distribution family when parametric, probability classes, interval level, uncertainty source, calibration method, calibration sample, scoring rules, held-out design, and any operations unavailable for the representation. A manuscript should not use “uncertainty interval” as a generic label when the method can be named more precisely.

## 21. Minimum probabilistic checklist

- [ ] predictive representation named;
- [ ] class probabilities normalized and named;
- [ ] uncertainty source and target stated;
- [ ] scores matched to the representation;
- [ ] final test labels excluded from calibration fitting;
- [ ] coverage paired with interval width or score;
- [ ] calibration paired with discrimination/ranking where relevant;
- [ ] no density claimed from interval-only output;
- [ ] no posterior interpretation assigned to bootstrap/ensemble variation;
- [ ] manifest preserves model, split, score, and calibration metadata.


### Extended scenario 1: Disease-risk probabilities

A crop disease classifier outputs infection probabilities for scouting decisions. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Evaluate log loss, Brier score, calibration, and an operational threshold separately. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is using accuracy as evidence that the numerical probabilities are trustworthy. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Show a reliability table and the number of observations contributing to each probability region. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 2: Yield prediction intervals

A model predicts yield with a central 90% interval. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Evaluate empirical coverage and width on design-appropriate held-out observations. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is reporting nominal 90% as if empirical coverage had already been demonstrated. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report misses, interval width, and whether spread changes across the predictor range. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 3: Heteroscedastic soil response

Measurement variability increases with the predicted soil property. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use a distributional or quantile representation that can change spread with predictors. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is adding one global residual standard deviation to every point prediction. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Compare interval widths across low- and high-variance regions. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 4: Multiclass nutrient deficiency

A classifier returns probabilities for several deficiency classes. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Score the full probability vector and inspect one-vs-rest calibration by class. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is renormalizing arbitrary scores and calling them calibrated probabilities. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report class prevalence and calibration gaps for rare classes. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 5: Ensemble prediction

Several fitted learners generate an empirical prediction distribution. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Treat ensemble samples as an empirical representation and state what sources of variation they contain. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is calling ensemble spread a Bayesian posterior standard deviation. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Compare ensemble spread with observed error and calibration. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 6: Bootstrap prediction

Repeated bootstrap fits produce a predictive sample per observation. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Label the uncertainty source as bootstrap and distinguish parameter/statistical resampling from outcome noise. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is interpreting bootstrap samples as direct draws from the posterior predictive distribution. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report bootstrap scheme, sampling unit, successful replicates, and interval construction. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 7: Quantile regression

A model provides 0.1, 0.5, and 0.9 conditional quantiles. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use quantile and interval operations while leaving density operations unavailable. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is inventing a Normal density solely from three quantiles. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Evaluate quantile coverage and crossing or ordering behavior. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 8: Climate probability forecast

A seasonal model predicts categories of dry, normal, and wet conditions. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use multiclass probability scores and calibration across truly future seasons. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is calibrating and evaluating on the same future years. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Separate calibration years from the final forecast-evaluation period. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 9: Disease-risk probabilities

A crop disease classifier outputs infection probabilities for scouting decisions. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Evaluate log loss, Brier score, calibration, and an operational threshold separately. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is using accuracy as evidence that the numerical probabilities are trustworthy. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Show a reliability table and the number of observations contributing to each probability region. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 10: Yield prediction intervals

A model predicts yield with a central 90% interval. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Evaluate empirical coverage and width on design-appropriate held-out observations. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is reporting nominal 90% as if empirical coverage had already been demonstrated. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report misses, interval width, and whether spread changes across the predictor range. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 11: Heteroscedastic soil response

Measurement variability increases with the predicted soil property. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use a distributional or quantile representation that can change spread with predictors. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is adding one global residual standard deviation to every point prediction. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Compare interval widths across low- and high-variance regions. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 12: Multiclass nutrient deficiency

A classifier returns probabilities for several deficiency classes. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Score the full probability vector and inspect one-vs-rest calibration by class. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is renormalizing arbitrary scores and calling them calibrated probabilities. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report class prevalence and calibration gaps for rare classes. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 13: Ensemble prediction

Several fitted learners generate an empirical prediction distribution. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Treat ensemble samples as an empirical representation and state what sources of variation they contain. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is calling ensemble spread a Bayesian posterior standard deviation. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Compare ensemble spread with observed error and calibration. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 14: Bootstrap prediction

Repeated bootstrap fits produce a predictive sample per observation. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Label the uncertainty source as bootstrap and distinguish parameter/statistical resampling from outcome noise. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is interpreting bootstrap samples as direct draws from the posterior predictive distribution. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report bootstrap scheme, sampling unit, successful replicates, and interval construction. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 15: Quantile regression

A model provides 0.1, 0.5, and 0.9 conditional quantiles. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use quantile and interval operations while leaving density operations unavailable. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is inventing a Normal density solely from three quantiles. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Evaluate quantile coverage and crossing or ordering behavior. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 16: Climate probability forecast

A seasonal model predicts categories of dry, normal, and wet conditions. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use multiclass probability scores and calibration across truly future seasons. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is calibrating and evaluating on the same future years. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Separate calibration years from the final forecast-evaluation period. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 17: Disease-risk probabilities

A crop disease classifier outputs infection probabilities for scouting decisions. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Evaluate log loss, Brier score, calibration, and an operational threshold separately. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is using accuracy as evidence that the numerical probabilities are trustworthy. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Show a reliability table and the number of observations contributing to each probability region. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 18: Yield prediction intervals

A model predicts yield with a central 90% interval. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Evaluate empirical coverage and width on design-appropriate held-out observations. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is reporting nominal 90% as if empirical coverage had already been demonstrated. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report misses, interval width, and whether spread changes across the predictor range. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 19: Heteroscedastic soil response

Measurement variability increases with the predicted soil property. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use a distributional or quantile representation that can change spread with predictors. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is adding one global residual standard deviation to every point prediction. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Compare interval widths across low- and high-variance regions. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 20: Multiclass nutrient deficiency

A classifier returns probabilities for several deficiency classes. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Score the full probability vector and inspect one-vs-rest calibration by class. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is renormalizing arbitrary scores and calling them calibrated probabilities. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report class prevalence and calibration gaps for rare classes. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 21: Ensemble prediction

Several fitted learners generate an empirical prediction distribution. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Treat ensemble samples as an empirical representation and state what sources of variation they contain. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is calling ensemble spread a Bayesian posterior standard deviation. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Compare ensemble spread with observed error and calibration. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 22: Bootstrap prediction

Repeated bootstrap fits produce a predictive sample per observation. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Label the uncertainty source as bootstrap and distinguish parameter/statistical resampling from outcome noise. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is interpreting bootstrap samples as direct draws from the posterior predictive distribution. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report bootstrap scheme, sampling unit, successful replicates, and interval construction. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 23: Quantile regression

A model provides 0.1, 0.5, and 0.9 conditional quantiles. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use quantile and interval operations while leaving density operations unavailable. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is inventing a Normal density solely from three quantiles. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Evaluate quantile coverage and crossing or ordering behavior. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 24: Climate probability forecast

A seasonal model predicts categories of dry, normal, and wet conditions. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use multiclass probability scores and calibration across truly future seasons. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is calibrating and evaluating on the same future years. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Separate calibration years from the final forecast-evaluation period. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 25: Disease-risk probabilities

A crop disease classifier outputs infection probabilities for scouting decisions. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Evaluate log loss, Brier score, calibration, and an operational threshold separately. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is using accuracy as evidence that the numerical probabilities are trustworthy. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Show a reliability table and the number of observations contributing to each probability region. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 26: Yield prediction intervals

A model predicts yield with a central 90% interval. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Evaluate empirical coverage and width on design-appropriate held-out observations. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is reporting nominal 90% as if empirical coverage had already been demonstrated. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report misses, interval width, and whether spread changes across the predictor range. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 27: Heteroscedastic soil response

Measurement variability increases with the predicted soil property. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use a distributional or quantile representation that can change spread with predictors. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is adding one global residual standard deviation to every point prediction. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Compare interval widths across low- and high-variance regions. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.
