---
title: "sciModelFlowR 0.3.0: Imbalance, Calibration, and Thresholds"
format:
  html:
    toc: true
    toc-depth: 3
execute:
  echo: true
  warning: true
  message: false
---

# Imbalance, calibration, and thresholds

Imbalanced classification is not one problem. It combines at least four distinct questions: how a learner sees the training data, how probabilities reflect the target population, how performance is summarized across classes, and how a probability is converted into an action. `sciModelFlowR` 0.3.0 keeps these questions separate.

The governing rule is:

**Sampling and weighting belong to analysis data. Calibration belongs to independent calibration evidence. Threshold choice belongs to a declared decision objective. Final test labels belong to evaluation only.**

## 1. Learning objectives

After this vignette, the reader should be able to explain why class imbalance does not automatically justify SMOTE, apply class weights or sampling only inside training folds, distinguish training prevalence from deployment prevalence, evaluate minority-class performance with appropriate metrics, recalibrate probabilities after imbalance handling, choose a threshold on validation/calibration data, and document the entire sequence without allowing final test labels to influence a decision rule.

## 2. The Gold multiclass fixture

```{r}
#| eval: false
library(sciModelFlowR)

d <- smf_load_dataset("gold_multiclass_imbalanced")
table(d$class)
```

Class `C` is deliberately uncommon. The `true_p_A`, `true_p_B`, and `true_p_C` columns are generator metadata and must be excluded from predictors. Their purpose is validation of simulated behavior, not model fitting.

The dataset is intentionally not perfectly separable. This prevents a sophisticated algorithm from making the imbalance problem disappear simply because classes are trivial to distinguish.

## 3. Declare imbalance handling

```{r}
#| eval: false
smf_imbalance_spec("none")
smf_imbalance_spec("weights")
smf_imbalance_spec("downsample")
smf_imbalance_spec("upsample")
smf_imbalance_spec("smote", ratio = 1, neighbors = 5L)
```

An `ImbalanceSpec` is part of the experiment specification. This makes the sampling rule auditable and prevents a notebook from applying an undocumented transformation before the package sees the data.

## 4. The fold boundary is non-negotiable

Oversampling before cross-validation allows synthetic or duplicated information derived from assessment observations to enter training. Even when the exact assessment row is not copied, nearest-neighbor synthesis can encode its geometry.

`sciModelFlowR` therefore refuses to apply imbalance learning with `context = "assessment"` or `context = "test"`. In resampling, the operation is invoked separately for each analysis fold after preprocessing and feature-state learning have established the training boundary.

## 5. Class weights

Inverse-frequency weights increase the contribution of observations from uncommon classes without fabricating new predictor vectors. They are often a useful first alternative when the selected learner supports case weights.

```{r}
#| eval: false
weighted <- smf_apply_imbalance(
  data = training_table,
  outcome = "class",
  spec = smf_imbalance_spec("weights"),
  context = "analysis"
)
```

Weights change the fitting objective. They do not automatically change the event prevalence represented by the raw data, and they do not guarantee calibrated probabilities for the target population.

## 6. Random undersampling

Undersampling can reduce majority-class dominance and computation but discards observed information. The retained majority examples also vary with the seed. Use it when there is a clear reason, then evaluate stability across repetitions rather than treating one sampled training set as uniquely correct.

If the majority class contains important substructure, aggressive undersampling can remove rare but scientifically relevant majority patterns.

## 7. Random oversampling

Random oversampling duplicates minority observations. It increases their contribution to the fitting process but does not create new independent information. A learner may overfit duplicated rare examples, particularly when the minority sample is very small.

The scientific unit remains the original unit. Replicated rows created for training must never be counted as additional experimental replication in inference or reporting.

## 8. SMOTE and variants

The optional `themis` adapter provides SMOTE-family operations. In the managed workflow, they are used only on analysis data. The synthetic observations are not carried into the assessment set, calibration set, or scientific sample-size accounting.

SMOTE requires a meaningful predictor geometry. Distance after scaling numeric predictors can be useful; distance across poorly encoded categories or irrelevant high-dimensional variables can be misleading. The existence of an implementation is not evidence that synthetic interpolation is scientifically appropriate.

## 9. Sampling order and preprocessing

A defensible sequence is generally:

1. define the analysis fold;
2. fit training-only preprocessing;
3. fit training-only feature state;
4. apply the resulting representation to the analysis rows;
5. learn weights or synthetic sampling only from those rows;
6. fit the model;
7. transform assessment rows with the saved preprocessing/feature state only;
8. predict assessment rows without resampling them.

This order keeps the assessment outcome invisible to the entire training pipeline.

## 10. Why accuracy is inadequate

A model that predicts the majority class for every observation can have high accuracy when prevalence is extreme. Balanced accuracy averages sensitivity across classes and therefore gives rare classes equal class-level importance. MCC summarizes the full binary confusion matrix and is often informative under imbalance. PR-AUC focuses attention on precision-recall behavior for a chosen event.

The right metric depends on the scientific cost structure. No imbalance method should be judged solely by how much it improves one metric selected after looking at the test set.

## 11. ROC-AUC versus PR-AUC

ROC-AUC summarizes ranking between positive and negative cases and is relatively insensitive to prevalence. This can be useful, but a high ROC-AUC can coexist with low precision when the event is rare. PR-AUC makes false positives more visible through precision.

For rare pest detection, disease screening, or failure prediction, show both ranking and operational consequences rather than using one curve as a universal performance certificate.

## 12. Calibration after reweighting or sampling

Changing the training class distribution can change the relationship between model scores and deployment probabilities. A classifier trained on artificially balanced data may rank cases well while its raw probabilities no longer reflect the target prevalence.

Version 0.3.0 therefore allows an explicit calibration stage using a subset of analysis-side data that was not used for model fitting. Final test labels remain protected.

## 13. Calibration specification

```{r}
#| eval: false
cal_spec <- smf_calibration_spec(
  method = "platt",
  source = "analysis_holdout",
  calibration_prop = 0.20,
  bins = 10L
)
```

The managed experiment subdivides its training portion into a model-fitting subset and a calibration subset. Preprocessing and feature state are learned on the model-fitting subset, then applied to calibration data. The calibration model is estimated from calibration predictions and outcomes. Only then is the untouched final test set predicted and calibrated.

## 14. Platt calibration

Platt calibration fits a logistic relationship between the model score and the observed binary outcome. In the package-native implementation, probabilities are converted to log odds before fitting the calibration model.

It is appropriate when a smooth monotonic correction is scientifically plausible and the calibration sample is sufficient. It cannot rescue a model that contains no useful ranking information.

## 15. Isotonic calibration

Isotonic regression estimates a monotonic but flexible probability mapping. It can represent nonlinear miscalibration without choosing a sigmoid form. The flexibility also means that small calibration samples can produce unstable step functions.

For multiclass outcomes, 0.3.0 applies one-vs-rest isotonic maps and renormalizes the resulting probabilities. The report should mention that this renormalization can affect perfect one-vs-rest calibration.

## 16. Multinomial calibration

Multiclass calibration can be modeled jointly through a multinomial calibration adapter. Version 0.3.0 provides an optional `nnet`-based path using log-probability contrasts as predictors. This is more coherent than applying a binary Platt model independently and pretending the outputs will necessarily sum to one.

Multinomial calibration requires enough observations from every class. If a rare class has only a handful of calibration examples, the correct response may be to report uncertainty and gather more data rather than to fit a high-dimensional calibration model.

## 17. Beta calibration status

The specification reserves `beta` calibration because the method is useful for probability distortions not captured by a single sigmoid. The 0.3.0 package does not automatically select it. The optional `probably`/`betacal` route remains adapter-gated until runtime validation is completed in the final local validation cycle.

A declared method that cannot be certified should fail explicitly rather than fall back to Platt calibration.

## 18. Reliability diagrams

`smf_calibration_report()` groups probability predictions into bins and reports mean predicted probability, observed frequency, sample size, and absolute gap. `smf_plot(..., type = "calibration")` can visualize the same relationship.

A reliability diagram should show how much data support each region. A point at 0.9 based on three observations is not equivalent to a point based on three hundred observations.

## 19. Expected calibration error

ECE is the weighted mean of absolute bin-wise calibration gaps. It is compact and useful for comparison, but it changes with the number and placement of bins. Avoid reporting ECE to several decimal places without saying how it was calculated.

Calibration plots, Brier score, log loss, and class counts provide necessary context.

## 20. Thresholds are decisions, not calibration

A calibrated probability does not specify the correct classification threshold. `smf_optimize_threshold()` searches a user-provided threshold grid using calibration or validation evidence and a declared metric.

```{r}
#| eval: false
thr <- smf_optimize_threshold(
  truth = validation_truth,
  probability = validation_probability,
  metric = "balanced_accuracy",
  positive_label = "event"
)
```

The chosen threshold belongs in the report and manifest. It must not be recomputed on the final test set.

## 21. Cost-sensitive decisions

Balanced accuracy, F1, or MCC are generic criteria. A real intervention may have an asymmetric loss: missing a severe pest outbreak can cost far more than an unnecessary scouting visit. In that case the scientifically meaningful decision rule should use explicit costs or utilities.

Version 0.3.0 provides metric-based threshold optimization as a foundation. A later decision-analysis layer can incorporate richer cost structures without changing the probability model itself.

## 22. Prevalence shift

Calibration is conditional on the population represented by the calibration data. If deployment prevalence changes, a previously calibrated model may no longer be calibrated. This is especially important for disease surveillance across seasons, regions, cultivars, or management systems.

External validation should therefore report event prevalence and probability calibration, not only discrimination.

## 23. Binary workflow

```{r}
#| eval: false
b <- smf_load_dataset("gold_binary_calibration")
b$event <- factor(ifelse(b$event == 1, "yes", "no"), levels = c("no", "yes"))

spec <- smf_experiment_spec(
  task = smf_task_spec("binary", "event", positive_label = "yes"),
  data = smf_data_spec("event", c("x1", "x2"), id_column = "obs_id"),
  design = smf_design_spec(id_column = "obs_id"),
  preprocessing = smf_preprocess_spec(center = TRUE, scale = TRUE),
  resampling = smf_resampling_spec("stratified_holdout", .75, strata = "event"),
  model = smf_model_spec("logistic_regression", "stats"),
  metrics = list(
    smf_metric_spec("balanced_accuracy"),
    smf_metric_spec("log_loss"),
    smf_metric_spec("brier")
  ),
  imbalance = smf_imbalance_spec("weights", threshold = .40),
  calibration = smf_calibration_spec("platt", calibration_prop = .20)
)

fit <- smf_fit_experiment(spec, b)
```

This sequence is deliberately explicit. Weight learning uses only the model-training portion. Platt calibration uses a separate calibration portion. The final test labels contribute only to the reported metrics.

## 24. Multiclass workflow

For the imbalanced multiclass Gold fixture, prefer class weights or fold-safe sampling coupled with a backend that can produce multiclass probabilities. Calibration can use isotonic one-vs-rest or a multinomial method when there is enough calibration evidence.

The final report should include the confusion matrix or class-wise recall, probability scores, class prevalence, and calibration by class. A pooled accuracy value hides too much information.

## 25. Scientifically inappropriate workflow

The following procedure is invalid: apply SMOTE to the complete dataset, optimize the probability threshold on the test set, fit isotonic calibration on those same test predictions, then report the improved test Brier score.

Every post-processing choice has used the labels that were supposed to evaluate generalization. The reported test result is therefore a development score, not independent evidence. A correct workflow reserves calibration/validation data for these decisions and keeps the final test set untouched until the entire pipeline is fixed.

## 26. Reporting checklist

- [ ] original class prevalence reported;
- [ ] scientific event/positive class defined;
- [ ] resampling unit declared;
- [ ] weighting or sampling method named;
- [ ] sampling performed only within analysis folds;
- [ ] synthetic rows excluded from scientific sample-size claims;
- [ ] calibration source declared;
- [ ] calibration sample separated from final test;
- [ ] calibration method and bins reported;
- [ ] threshold selection data and criterion reported;
- [ ] balanced/class-specific metrics shown;
- [ ] log loss or Brier score shown when probabilities matter;
- [ ] external/deployment prevalence discussed;
- [ ] manifest records sampling, calibration, threshold, seed, and backend.


### Extended scenario 1: Pest outbreak detection

Only a small fraction of scouting observations contain the target pest. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use group-aware folds if multiple observations come from the same field and learn weights or sampling inside each analysis fold. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is oversampling before the field split. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report precision, recall, PR-AUC, probability calibration, and the action threshold. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 2: Disease screening

A sensor model identifies diseased plants when disease prevalence varies by greenhouse. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Treat greenhouse as a grouping or external domain and examine prevalence-specific calibration. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is using a single random split that mixes greenhouse effects. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report calibration and sensitivity by greenhouse as well as pooled performance. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 3: Rare nutrient disorder

A multiclass image model has one rare deficiency category. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Preserve class labels in every probability matrix and use class-wise recall plus multiclass log loss. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is optimizing pooled accuracy while never detecting the rare disorder. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Show the rare-class sample count and its calibration evidence. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 4: Quality rejection

A classifier flags rare unacceptable produce where false negatives are expensive. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Choose the threshold using a validation objective that reflects the asymmetric consequence. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is choosing 0.5 because it is conventional. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report the decision criterion and sensitivity analysis across plausible thresholds. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 5: Seasonal prevalence shift

A disease event is common in wet seasons and uncommon in dry seasons. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use temporally honest validation and evaluate calibration separately by season. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is assuming one calibration map remains valid under changed base rates. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report seasonal prevalence, calibration gap, and threshold implications. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 6: Spatially clustered rare events

Rare soil contamination cases occur in spatial clusters. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use spatial validation and ensure synthetic sampling never crosses the assessment boundary. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is random row-wise SMOTE that creates near-neighbors of held-out locations. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report spatial separation and class-specific errors. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 7: Instrument anomaly detection

Failures are rare and instruments contribute repeated records. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Keep instrument identity grouped and evaluate thresholds on unseen instruments when deployment requires it. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is letting records from the same failing instrument appear on both sides of a split. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report instrument-level event counts and external calibration. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 8: Breeding selection

A rare superior genotype class is predicted across environments. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Separate environment transfer from class imbalance and do not create synthetic genotypes that are interpreted as real experimental units. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is counting oversampled records as biological replication. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report original genotype counts and environment-wise probability performance. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 9: Pest outbreak detection

Only a small fraction of scouting observations contain the target pest. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use group-aware folds if multiple observations come from the same field and learn weights or sampling inside each analysis fold. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is oversampling before the field split. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report precision, recall, PR-AUC, probability calibration, and the action threshold. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 10: Disease screening

A sensor model identifies diseased plants when disease prevalence varies by greenhouse. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Treat greenhouse as a grouping or external domain and examine prevalence-specific calibration. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is using a single random split that mixes greenhouse effects. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report calibration and sensitivity by greenhouse as well as pooled performance. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 11: Rare nutrient disorder

A multiclass image model has one rare deficiency category. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Preserve class labels in every probability matrix and use class-wise recall plus multiclass log loss. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is optimizing pooled accuracy while never detecting the rare disorder. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Show the rare-class sample count and its calibration evidence. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 12: Quality rejection

A classifier flags rare unacceptable produce where false negatives are expensive. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Choose the threshold using a validation objective that reflects the asymmetric consequence. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is choosing 0.5 because it is conventional. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report the decision criterion and sensitivity analysis across plausible thresholds. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 13: Seasonal prevalence shift

A disease event is common in wet seasons and uncommon in dry seasons. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use temporally honest validation and evaluate calibration separately by season. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is assuming one calibration map remains valid under changed base rates. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report seasonal prevalence, calibration gap, and threshold implications. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 14: Spatially clustered rare events

Rare soil contamination cases occur in spatial clusters. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use spatial validation and ensure synthetic sampling never crosses the assessment boundary. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is random row-wise SMOTE that creates near-neighbors of held-out locations. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report spatial separation and class-specific errors. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 15: Instrument anomaly detection

Failures are rare and instruments contribute repeated records. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Keep instrument identity grouped and evaluate thresholds on unseen instruments when deployment requires it. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is letting records from the same failing instrument appear on both sides of a split. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report instrument-level event counts and external calibration. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 16: Breeding selection

A rare superior genotype class is predicted across environments. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Separate environment transfer from class imbalance and do not create synthetic genotypes that are interpreted as real experimental units. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is counting oversampled records as biological replication. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report original genotype counts and environment-wise probability performance. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 17: Pest outbreak detection

Only a small fraction of scouting observations contain the target pest. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use group-aware folds if multiple observations come from the same field and learn weights or sampling inside each analysis fold. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is oversampling before the field split. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report precision, recall, PR-AUC, probability calibration, and the action threshold. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 18: Disease screening

A sensor model identifies diseased plants when disease prevalence varies by greenhouse. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Treat greenhouse as a grouping or external domain and examine prevalence-specific calibration. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is using a single random split that mixes greenhouse effects. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report calibration and sensitivity by greenhouse as well as pooled performance. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 19: Rare nutrient disorder

A multiclass image model has one rare deficiency category. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Preserve class labels in every probability matrix and use class-wise recall plus multiclass log loss. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is optimizing pooled accuracy while never detecting the rare disorder. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Show the rare-class sample count and its calibration evidence. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 20: Quality rejection

A classifier flags rare unacceptable produce where false negatives are expensive. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Choose the threshold using a validation objective that reflects the asymmetric consequence. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is choosing 0.5 because it is conventional. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report the decision criterion and sensitivity analysis across plausible thresholds. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 21: Seasonal prevalence shift

A disease event is common in wet seasons and uncommon in dry seasons. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use temporally honest validation and evaluate calibration separately by season. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is assuming one calibration map remains valid under changed base rates. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report seasonal prevalence, calibration gap, and threshold implications. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 22: Spatially clustered rare events

Rare soil contamination cases occur in spatial clusters. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use spatial validation and ensure synthetic sampling never crosses the assessment boundary. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is random row-wise SMOTE that creates near-neighbors of held-out locations. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report spatial separation and class-specific errors. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 23: Instrument anomaly detection

Failures are rare and instruments contribute repeated records. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Keep instrument identity grouped and evaluate thresholds on unseen instruments when deployment requires it. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is letting records from the same failing instrument appear on both sides of a split. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report instrument-level event counts and external calibration. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 24: Breeding selection

A rare superior genotype class is predicted across environments. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Separate environment transfer from class imbalance and do not create synthetic genotypes that are interpreted as real experimental units. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is counting oversampled records as biological replication. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report original genotype counts and environment-wise probability performance. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.


### Extended scenario 25: Pest outbreak detection

Only a small fraction of scouting observations contain the target pest. In a `sciModelFlowR` analysis, the first step is to translate this situation into an explicit prediction target and a validation unit. Use group-aware folds if multiple observations come from the same field and learn weights or sampling inside each analysis fold. The important point is that the estimator is not allowed to redefine the scientific question after performance results are visible.

The main failure mode is oversampling before the field split. A defensible workflow stores the split or resampling geometry, fits all learned transformations only on analysis rows, and evaluates the relevant quantities on held-out observations. Report precision, recall, PR-AUC, probability calibration, and the action threshold. The evidence should be reported together with the model family, backend, probability or response scale, warnings, seed, and manifest hash. This keeps the result interpretable even if a different backend is evaluated later.

## 1.0.0 consolidation note

In the 1.0.0 Consolidated Scientific Release, this workflow keeps the same scientific semantics established during development. The public function contract is frozen from 0.9.0; final certification changes evidence and backend status, not the design, leakage, uncertainty, or provenance rules taught in this vignette.

